HO unification from object language to meta language

### 2024-02-15 main.tex
wip
Update tex

@@

\def\ol{\acronym{ol}}
\def\ml{\acronym{ml}}

@@

  Unification is a fundamental process behind goal resolution in logic
  programming. Two terms $t_1$ and $t_2$ unify up to $\eta\beta$-reduction if it
  exist a substitution $\theta$ such that $\theta t_1 = \theta t_2$. A
  substitution is a mapping from unification variables to terms and it is the most
  general, if forall substitution $\theta'$, such that $\theta' t_1 = \theta'
  t_2$, it exists a substitution $\theta''$ such that $\theta' = \theta \theta''$.
  Unification has been widely studied in the 90s [here the refs]. In this paper we
  want to give a framework to reason about unification in meta-programming.

@@

@@

Meta programming \cite{hill1994} is a programming technique in which a program
can treat an other program as its data. This latter program is called object
language (\ol for short), while the former is called meta language (\ml for
short). At the heart of meta programming lays the necessity of representing
terms of the \ol in the \ml so that a wide set of program
manipulations ranging from interpretation to compilation. Furthermore, since a
logic program is a set of clauses, the behavior of the \ol can be
modified \texttt{in corsa}\todo{translate in english} by simply adding new rules inside the database of the meta program.

@@

\begin{tabular}{c}
\begin{minipage}{.9\textwidth}
\begin{minted}{elpi}
% deref
equal S (uv N) T1 :- assigned? N S T, equal S T T1.
equal S T1 (uv N) :- assigned? N S T, equal S T1 T.
equal S (app [uv N|Args]) T1 :- assigned? N S F, beta F Args T, equal S T T1.
equal S T1 (app [uv N|Args]) :- assigned? N S F, beta F Args T, equal S T1 T.
% congruence
equal S (app L1) (app L2) :- forall2 (equal S) L1 L2.
equal S (lam F1) (lam F2) :- pi x\ equal S x x => equal S (F1 x) (F2 x).
equal _ (c X) (c X).
equal _ (uv N) (uv N).
% beta
equal S (app [lam X | TL]) T :- beta (lam X) TL T', equal S T' T.
equal S T (app [lam X | TL]) :- beta (lam X) TL T', equal S T T'.
% eta
equal S (lam F) T :- not (T = lam _),
  pi x\ beta T [x] (T' x), equal S (lam F) (lam T').
equal S T (lam F) :- not (T = lam _),
  pi x\ beta T [x] (T' x), equal S (lam T') (lam F).
\end{minted}
\end{minipage}
\\\hline
\begin{minipage}{.9\textwidth}
\begin{minted}{elpi}
% deref
equal S (uv N Args) T1 :- assigned? N S F, beta-abs F Args T, equal S T T1.
equal S T1 (uv N Args) :- assigned? N S F, beta-abs F Args T, equal S T1 T.
% congruence
equal S (app L1) (app L2) :- forall2 (equal S) L1 L2.
equal S (lam F1) (lam F2) :- pi x\ equal S x x => equal S (F1 x) (F2 x).
equal _ (c X) (c X).
equal _ (uv N A1) (uv N A2) :- forall2 (equal S) A1 A2. %fixme in the code
\end{minted}
\end{minipage}

@@

\section{recoving binary app}

>>>>

\newcommand{\library}[1]{\textit{#1}\xspace}
\def\stdpp{\library{stdpp}}
\def\iris{\library{iris}}
\tableofcontents

@@

\def\ol{\acronym{ol}} % object language
\def\ml{\acronym{ml}} % meta language
\def\lf{\acronym{lf}} % logical framework

@@

  % Unification is a fundamental process behind goal resolution in logic
  % programming. Two terms $t_1$ and $t_2$ unify up to $\eta\beta$-reduction if it
  % exist a substitution $\theta$ such that $\theta t_1 = \theta t_2$. A
  % substitution is a mapping from unification variables to terms and it is the most
  % general, if forall substitution $\theta'$, such that $\theta' t_1 = \theta'
  % t_2$, it exists a substitution $\theta''$ such that $\theta' = \theta \theta''$.
  % Unification has been widely studied in the 90s [here the refs]. In this paper we
  % want to give a framework to reason about unification in meta-programming.
  Meta-programming \cite{hill1994} is a programming technique in which a
  program, referred to as meta language (\ml), can treat an other program, knwon
  as the object language (\ol) as its data. This programming approach is used in
  various works \cite{cite1, cite2...} to construct logical frameworks (\lf)
  capable of automating proof resolution. In many of the previously the
  previously mentioned works, the equational theory expressed by the \ol is
  weaker than the \ml one. Consequently, no effort is required to adjust the
  unification algorithm of the \ml since it encompass the \ml one. In our
  specific scenario, however, we work with an \ol with a richer equational
  theory than our \ml. This paper introduces a framework that enable the \ml to
  comprehend the unification of the \ol and provides some practical applications
  of these findings. Additionally, we extend our work to unification in the case
  of higher-order variable quantification.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Context
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

% meta programming
Meta programming is a powerful programming style that enable the incorporation
of an object language (\ol) into a meta language (\ml). This control over the
\ol lies at the core of meta-programming, facilitating a spectrum of operations
on the object language. Some of the principal program manipulations range
from interpretation to compilation.

% theorem prover and tc
Thanks to meta programming, it is feasible to express the equational theory of a
theorem prover, as noted in \cite{citeX, Y Z}, with the \ml serving as the
foundational infrastructure for this ecosystem. Proving a theorem, that is
solving a query using the jargon of logic programming consist in verifying if a
proposition is a logical consequence of a given set of assumption. The benefit
of of using a prologish verification style is that the proof resolution can be
automatized thanks to different search strategies, such as a depth first search
with backtracking. One of the main expression of this proof automation is
type-class resolution\cite{wadler89, sozeau08}. Type classes are a typing
structure to introduce ad-hoc polymorphism in functional programming languages.
In \coq, the use of type class has become a more and more used programming
style. Several are the libraries based on this machinery, we can site \stdpp and
\iris.

% embed tc in ml
One of the

\begin{itemize}
  \item about meta programming (what it is, when to use it)
  \item theorem provers (proof automation)
\end{itemize}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Example
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\begin{itemize}
  \item TODO: context
  \item TODO: example
  \item TODO: riassunto delle next sections
\end{itemize}

@@

Meta programming \cite{hill1994} is a programming technique in which a program
can treat an other program as its data. This latter program is called object
language (\ol for short), while the former is called meta language (\ml for
short). At the heart of meta programming lays the necessity of representing
terms of the \ol in the \ml so that a wide set of program
manipulations ranging from interpretation to compilation. 
% Furthermore, since a
% logic program is a set of clauses, the behavior of the \ol can be
% modified \texttt{in corsa}\todo{translate in english} by simply adding new rules inside the database of the meta program.

@@

\noindent
\begin{tabular}{c}
\begin{minipage}{.95\textwidth}
\begin{minted}{elpi}
% deref
equal S (uv N) T1 :- assigned? N S T, equal S T T1.
equal S T1 (uv N) :- assigned? N S T, equal S T1 T.
equal S (app [uv N|A]) T1 :- assigned? N S F, beta F A T, equal S T T1.
equal S T1 (app [uv N|A]) :- assigned? N S F, beta F A T, equal S T1 T.
% congruence
equal S (app L1) (app L2) :- forall2 (equal S) L1 L2.
equal S (lam F1) (lam F2) :- pi x\ equal S x x => equal S (F1 x) (F2 x).
equal _ (c X) (c X).
equal _ (uv N) (uv N).
% beta
equal S (app [lam X | TL]) T :- beta (lam X) TL T', equal S T' T.
equal S T (app [lam X | TL]) :- beta (lam X) TL T', equal S T T'.
% eta
equal S (lam F) T :- not (T = lam _),
  pi x\ beta T [x] (T' x), equal S (lam F) (lam T').
equal S T (lam F) :- not (T = lam _),
  pi x\ beta T [x] (T' x), equal S (lam T') (lam F).
\end{minted}
\end{minipage}
\\\hline
\begin{minipage}{.95\textwidth}
\begin{minted}{elpi}
% deref
equal S (uv N A) T1 :- assigned? N S F, beta-abs F A T, equal S T T1.
equal S T1 (uv N A) :- assigned? N S F, beta-abs F A T, equal S T1 T.
% congruence
equal S (app L1) (app L2) :- forall2 (equal S) L1 L2.
equal S (lam F1) (lam F2) :- pi x\ equal S x x => equal S (F1 x) (F2 x).
equal _ (c X) (c X).
equal _ (uv N A1) (uv N A2) :- forall2 (equal S) A1 A2. %fixme in the code
\end{minted}
\end{minipage}

@@

\section{heuristic / binary app}
